Generalized canonical rings of $\bq$-divisors on minimal rational surfaces

### 2017-03-24 surfaces-write-up/surface-paper.tex
added edits for second responses file

\subsection{Notation}
We now define some notation used throughout the paper.
Use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ and the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let 
$R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the generalized canonical ring associated to the $\bq$-divisor $D$. 
So, $R_D$ is notation for $R(X,D)$ when $X$ is understood from context.
If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. 
Throughout, we work over a fixed algebraically closed field $\bk$.
Since generator and relation degrees are preserved under base change to the algebraic closure, assuming that $\bk$ is algebraically closed is no restriction
on generality and the results hold equally well over an arbitrary field.
If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

\begin{example}
\label{eg:radical}
Consider $\frac{-1}{5}V(x_0^2 + x_1^2 + x_2^2) + \frac{1}{7}V(x_0^2 + x_1^2 +
x_3^2) + \frac{1}{17}V(x_0^2 + x_2^2 + x_3^2) - \frac{1}{596}V(x_1^2 + x_2^2 +
x_3^2)$. In degree $355216 = 5 \cdot 7 \cdot 17 \cdot 596$, this has dimension
$6$.
However, all lower degrees only have
dimension at most 1. Therefore, we cannot hope to bound the degree
of generation of $R_D$ as a linear combination of $\ell_i$,
in analogy to \cite[Theorem 8]{dorney:canonical}
unless we require that $D$ includes ghost divisors. That is, unless
$D_0,\ldots D_m$ are taken to be linearly independent hyperplanes.
\end{example}

@@

One way to describe the Hirzebruch surface $\hirz_m$ is as a
quotient,
\[
\hirz_m \cong (\ba^2 \setminus \{0\}) \times (\ba^2 \setminus \{0\})/\mathbb G_m \times \mathbb G_m
\]
where $\mathbb G_m$ is the multiplicative subgroup of $\ba^1$, and the action of $\mathbb G_m \times \mathbb G_m$ is given by
$(\lambda, \mu) \cdot (u\colon v; z\colon w) \mapsto (\lambda u\colon \lambda v; \mu z\colon \lambda^{-m} \mu w),$ as described in \cite[p.~ 6]{zhao:counting-cubic}. Hence, one can think of
$\hirz_m$ as a $\bp^1$ bundle where $u,v$ are the coordinates on
$\bp^1$ and $z,w$ are the coordinates on the fiber.
Sections of a line bundle $\mathscr L$ on $\hirz_m$ can be written as
rational functions in $z, w, u, v$.
Furthermore we define the bi-degree of a monomial $u^a v^{b} z^c w^d$
on $\hirz_n$ to be
$(a + b + mc, c + d)$. Rational sections of $\hirz_m$ can be written as
rational functions with numerators and denominators with the same bi-degrees. To see this, 
observe
$\Pic(\hirz_m) \cong \bz \times \bz$, where the class of a line bundle
in $\Pic(\hirz_m)$
is determined by its bi-degree, as follows from the excision exact sequence
for class groups.

>>>>

\subsection{Notation}
We now define some notation used throughout the paper.
We use $m \in \bz$ to index the dimension of a given projective
space $\bp^m$ and the type of the Hirzebruch surface $\hirz_m$.
Let $D$ be a (Weil) $\bq$-divisor of the form
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \bq \otimes_\bz \di X.
\end{align*}\noindent
where $n \in \bz$ indexes the number of irreducible divisors in 
the above expansion of $D$, $\alpha_i \in \bq$, and $D_i \in \di X$ is an integral
codimension 1 closed subscheme of $X$. When it is convenient to do
so, we shall sometimes start the indexing at $0$, so that $i$ runs
from $0$ to $n$. In the case $X = \bp^m$, define the degree of $D$ by $\deg D :=
\sum_{i=1}^{n}\alpha_i \cdot \deg D_i$. The floor of a $\bq$-divisor $D$ is a divisor $\lfloor D
\rfloor \in \di X$ is defined to be $\lfloor D \rfloor := \sum_{i = 1}^{n}
\lfloor \alpha_i \rfloor D_i$. Let 
$R_D := \bigoplus_{d \geq 0} u^d H^0(\bp^m, \lfloor dD \rfloor)$
denote the generalized canonical ring associated to the $\bq$-divisor $D$. 
We use $R_D$ as notation for $R(X,D)$ when $X$ is understood from context.
If $R$
is a graded ring, we denote the $d$th graded component of $R$ by $R_d$. If $r$ is a rational number, we let
$\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. 
Throughout, we work over a fixed algebraically closed field $\bk$.
Since generator and relation degrees are preserved under base change to the algebraic closure, assuming that $\bk$ is algebraically closed is no restriction
on generality and the results hold equally well over an arbitrary field.
If $D \in \bq \otimes_\bz
\di X$ is an arbitrary divisor, we denote $h^0(X, D) := \dim_\bk H^0(X,D)$.

@@

\begin{example}
\label{eg:radical}
Consider $D :=\frac{-1}{5}V(x_0^2 + x_1^2 + x_2^2) + \frac{1}{7}V(x_0^2 + x_1^2 +
x_3^2) + \frac{1}{17}V(x_0^2 + x_2^2 + x_3^2) - \frac{1}{596}V(x_1^2 + x_2^2 +
x_3^2)$. In degree $355216 = 5 \cdot 7 \cdot 17 \cdot 596$, $R_D$ has dimension
$6$.
However, for all $d \leq 355216,$ $h^0(\bp^3, dD) \leq 1$
and 
$h^0(\bp^3, dD)  = 1$ precisely when $\lfloor d D \rfloor = 0$ 
(so in this case, $H^0(\bp^3, d D)$ corresponds to the constant functions).
Therefore $R_D$ has a generator in degree $355216$.
Hence, we cannot hope to bound the degree
of generation of $R_D$ as a linear combination of $\ell_i$,
in analogy to \cite[Theorem 8]{dorney:canonical}
unless we require that $D$ includes ghost divisors. That is, unless
$D_0,\ldots D_m$ are taken to be linearly independent hyperplanes.
\end{example}

@@

One way to describe the Hirzebruch surface $\hirz_m$ is as a
quotient,
\[
\hirz_m \cong (\ba^2 \setminus \{0\}) \times (\ba^2 \setminus \{0\})/\mathbb G_m \times \mathbb G_m
\]
where $\mathbb G_m$ is the multiplicative subgroup of $\ba^1$, and the action of $\mathbb G_m \times \mathbb G_m$ is given by
$(\lambda, \mu) \cdot (u\colon v; z\colon w) \mapsto (\lambda u\colon \lambda v; \mu z\colon \lambda^{-m} \mu w),$ as described in \cite[p.~ 6]{zhao:counting-cubic}. Hence, one can think of
$\hirz_m$ as a $\bp^1$ bundle where $u,v$ are the coordinates on
$\bp^1$ and $z,w$ are the coordinates on the fiber.
Sections of a line bundle $\mathscr L$ on $\hirz_m$ can be written as
rational functions in $z, w, u, v$.
Furthermore we define the bi-degree of a monomial $u^a v^{b} z^c w^d$
on $\hirz_n$ to be
$(a + b + mc, c + d)$. Rational sections of $\hirz_m$ can be written as
rational functions with numerators and denominators of the same bi-degrees. To see this, 
observe
$\Pic(\hirz_m) \cong \bz \times \bz$, where the class of a line bundle
in $\Pic(\hirz_m)$
is determined by its bi-degree, as follows from the excision exact sequence
for class groups.
